Does an earthquake “know” how big it will be? A neural-net aided study

### 2024-06-24 article.tex
First draft of discussion.

Finally, Figure~\ref{fig:model_output}b presents the marginal probability density function (PDF) of magnitudes for MAGNET's prediction, $P(m)=N^{-1}\sum_i p_{\textbf{x}_i, t_i}(m)$. This represents the average probability distribution of magnitudes over the whole test set. It is seen that although the model was not constrained to do so, the average distribution aligns well with the ``uninformed'' GR distribution. That is, although MAGNET's prediction for each individual earthquake may deviate significantly from the GR distribution, it does reproduce this well known long-time behavior on average. It is also seen that the average prediction of the 300-day moving window GR deviates significantly from thte GR distribution. We also observe that MAGNET's average prediction aligns well with the \emph{test set} GR distribution, rather than the training set, suggesting its ability to generalize beyond the training data. This observation is noticeable even more so in the New Zealand and Japan data sets presented in Extended Data Fig. \ref{fig:model_output_em}.

@@

\begin{table}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/main_text_table.pdf}
    \caption{Mean score, $\mathcal{L}$, for various tested benchmarks. $\mathcal{L}$ is computed by Eq. \ref{eq:likelihood}. Lower score indicates a better magnitude predictor, best score in column is indicated in bold. First 3 columns display scores for the raw calculation of $\mathcal{L}$, middle and right column triplets display the scores for the temporally and spatially conditioned $\mathcal{L}$ scores, respectively.
    }
    \label{tab:mean_ll_benchmarks_main_text}
\end{table}

@@

\section{Discussion}
\neri{YOHAI LET'S MEET TO WRITE CMON}

@@

\newpage
\begin{table}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/em_table.pdf}
    \caption{Mean score, $\mathcal{L}$, for all tested benchmarks. $\mathcal{L}$ is computed by Eq. \ref{eq:likelihood}. Lower score indicates a better magnitude predictor, best score in column is indicated in bold. First 3 columns display scores for the raw calculation of $\mathcal{L}$, middle and right column triplets display the scores for the temporally and spatially conditioned $\mathcal{L}$ scores, respectively.
    }
    \label{tab:mean_ll_all_benchmarks}
\end{table}

>>>>

Finally, Figure~\ref{fig:model_output}b presents the marginal probability density function (PDF) of magnitudes for MAGNET's prediction, $P(m)=N^{-1}\sum_i p_{\textbf{x}_i, t_i}(m)$. This represents the average probability distribution of magnitudes over the whole test set. It is seen that although the model was not constrained to do so, the average distribution aligns well with the ``uninformed'' GR distribution. That is, although MAGNET's prediction for each individual earthquake may deviate significantly from the GR distribution, it does reproduce this well known long-time behavior on average. It is also seen that the average prediction of the 300-event moving window GR deviates significantly from the GR distribution. We also observe that MAGNET's average prediction aligns well with the \emph{test set} GR distribution, rather than the training set, suggesting its ability to generalize beyond the training data. This observation is noticeable even more so in the New Zealand and Japan data sets presented in Extended Data Fig. \ref{fig:model_output_em}.

@@

\begin{table}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/score_tables_main.pdf}
    \caption{Mean score, $\mathcal{L}$, for various tested benchmarks. $\mathcal{L}$ is computed by Eq. \ref{eq:likelihood}. Lower score indicates a better magnitude predictor, best score in column is indicated in bold. First 3 columns display scores for the raw calculation of $\mathcal{L}$, middle and right column triplets display the scores for the temporally and spatially conditioned $\mathcal{L}$ scores, respectively.
    }
    \label{tab:mean_ll_benchmarks_main_text}
\end{table}

@@

\section{Discussion / Implications of a successful magnitude predictor}
In this work we have introduced the concept of a \textit{magnitude predictor} to assist us in finding dependencies between magnitudes and the seismic history.  By focusing solely on magnitude prediction, assuming known location and timing of the event, we have shown the some of the information about an earthquakes magnitude is contained in the seismic history of a simple hypocenter catalog. Throughout this work we have made significant effort to address possible origins of this information gain. We have ruled out temporal and spatial variability in the completeness magnitude - and artifact originating from the seismic network's limitations. Furthermore, we examine a several benchmark models to make sure the advantage of our model, MAGNET, does not originate from a simple, previously attempted, statistical pattern of earthquake occurrence. Among the benchmarks we examine are standard spatially and temporally varying Gutenberg Richter distributions, as well as other benchmarks encompassing the near past seismic trends, such as a kernel density estimation of the past $n$ events. We have demonstrated that MAGNET's performance exceeds that of every benchmark tested, under any examined conditioning \neri{There is that temporal conditioning of the spatial GR I'm not sure how to treat in this sentence...}. It is then implied that MAGNET extracts previously unknown patterns of embodied in the seismic history that hold information of the magnitude of the upcoming earthquake.
A corollary follows this finding: for some earthquakes, the magnitude might be partially predetermined before the earthquake is initiated. This suggests the existence of a precursory process influencing an earthquake's magnitude. However, the specific nature and seismic patterns associated with this precursory stage remain unclear based on our current evidence. Nonetheless, this finding provides a promising avenue for future research aimed at identifying these precursory signals.
Investigating the properties of such precursory signals may be carried out by interpretability analysis of neural networks \neri{cite}. Such method may reveal what functions of the encoders (see architecture in methods section) have a larger weight in the result of well predicted events. Furthermore, the method of ablation \neri{cite} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.

\section{Possible affects on operational earthquake forecasting and prediction}
Discovery of specific pattern in the seismic history as precursory signals can advance the effort of earthquake prediction a great deal.\neri{cite? what?}. Nevertheless, discovering such signals in enough accuracy to prove useful for operational systems may be a further down the road.
An immediate application of the model presented in this work may be of quick magnitude estimation to be used for early warning systems. Such a system does not require additional data sources of assumptions on the model as the timing and location of the earthquake to be estimated are given by design of the task. 
A second application may incorporating such a magnitude predictor into an earthquake rate forecasting model such as the epidemic type aftershock sequence model (ETAS) \neri{cite}. In ETAS, the rate of events is determined by the product of a set of functions. \neri{formalize?} Origin time of events of the forecasted sequence are then sampled from the rate function, and assigned with a magnitude by sampling from a fitted GR distribution. y replacing the randomly assigned magnitudes of ETAS with MAGNET or a similar magnitude predictor one will improve the predictive power of rate prediction. As we have shown - magnitudes predicted by MAGNET will estimate the true magnitudes in a higher likelihood.

@@

\newpage
\begin{table}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/score_tables_EM.pdf}
    \caption{Mean score, $\mathcal{L}$, for all tested benchmarks. $\mathcal{L}$ is computed by Eq. \ref{eq:likelihood}. Lower score indicates a better magnitude predictor, best score in column is indicated in bold. First 3 columns display scores for the raw calculation of $\mathcal{L}$, middle and right column triplets display the scores for the temporally and spatially conditioned $\mathcal{L}$ scores, respectively.
    }
    \label{tab:mean_ll_all_benchmarks}
\end{table}
